Quantifying Limits to Detection of Early Warning for Critical Transitions

### 2012-01-13 boettiger.tex
rewrites..

%        File: boettiger.tex
%     Created: Mon Oct 10 04:00 PM 2011 P
% Last Change: Mon Oct 10 04:00 PM 2011 P
%%% PNAStmpl.tex
%% Template file to use for PNAS articles prepared in LaTeX
%% Version: Apr 14, 2008

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% BASIC CLASS FILE 
%% PNAStwo for two column articles is called by default.
%% Uncomment PNASone for single column articles. One column class
%% and style files are available upon request from pnas@nas.edu.
%% (uncomment means get rid of the '%' in front of the command)

%\documentclass{pnasone}
\documentclass{pnastwo}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Changing position of text on physical page:
%% Since not all printers position
%% the printed page in the same place on the physical page,
%% you can change the position yourself here, if you need to:

% \advance\voffset -.5in % Minus dimension will raise the printed page on the 
                         %  physical page; positive dimension will lower it.

                         %% You may set the dimension to the size that you need.

                         %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
                         %% OPTIONAL GRAPHICS STYLE FILE

                         %% Requires graphics style file (graphicx.sty), used for inserting
                         %% .eps files into LaTeX articles.
                         %% Note that inclusion of .eps files is for your reference only;
                         %% when submitting to PNAS please submit figures separately.

                         %% Type into the square brackets the name of the driver program 
                         %% that you are using. If you don't know, try dvips, which is the
                         %% most common PC driver, or textures for the Mac. These are the options:

                         % [dvips], [xdvi], [dvipdf], [dvipdfm], [dvipdfmx], [pdftex], [dvipsone],
                         % [dviwindo], [emtex], [dviwin], [pctexps], [pctexwin], [pctexhp], [pctex32],
                         % [truetex], [tcidvi], [vtex], [oztex], [textures], [xetex]

                         \usepackage[pdftex]{graphicx}

                         %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
                         %% OPTIONAL POSTSCRIPT FONT FILES

                         %% PostScript font files: You may need to edit the PNASoneF.sty
                         %% or PNAStwoF.sty file to make the font names match those on your system. 
                         %% Alternatively, you can leave the font style file commands commented out
                         %% and typeset your article using the default Computer Modern 
                         %% fonts (recommended). If accepted, your article will be typeset
                         %% at PNAS using PostScript fonts.

                         % Choose PNASoneF for one column; PNAStwoF for two column:
                         %\usepackage{PNASoneF}
                         %\usepackage{pnastwof}

                         %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
                         %% ADDITIONAL OPTIONAL STYLE FILES

                         %% The AMS math files are commonly used to gain access to useful features
                         %% like extended math fonts and math commands.

                         \usepackage{amssymb,amsfonts,amsmath}
                         %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
                         %% OPTIONAL MACRO FILES
                         %% Insert self-defined macros here.
                         %% \newcommand definitions are recommended; \def definitions are supported

                         %\newcommand{\mfrac}[2]{\frac{\displaystyle #1}{\displaystyle #2}}
                         %\def\s{\sigma}

                         \newcommand{\ud}{\mathrm{d}}
                         \renewcommand{\thefootnote}{\fnsymbol{footnote}}
                         %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
                         %% Don't type in anything in the following section:
                         %%%%%%%%%%%%
                         %% For PNAS Only:
                         \contributor{Submitted to Proceedings
                         of the National Academy of Sciences of the United States of America}
                         \url{www.pnas.org/cgi/doi/10.1073/pnas.0709640104}
                         \copyrightyear{2008}
                         \issuedate{Issue Date}
                         \volume{Volume}
                         \issuenumber{Issue Number}
                         %%%%%%%%%%%%

                         \begin{document}

                         %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

                         %% For titles, only capitalize the first letter
                         %% \title{Almost sharp fronts for the surface quasi-geostrophic equation}

                         \title{Quantifying Limits to Detection of Early Warning for Critical Transitions}

                         %% Enter authors via the \author command.  
                         %% Use \affil to define affiliations.
                         %% (Leave no spaces between author name and \affil command)

                         %% Note that the \thanks{} command has been disabled in favor of
                         %% a generic, reserved space for PNAS publication footnotes.

                         %% \author{<author name>
                         %% \affil{<number>}{<Institution>}} One number for each institution.
                         %% The same number should be used for authors that
                         %% are affiliated with the same institution, after the first time
                         %% only the number is needed, ie, \affil{number}{text}, \affil{number}{}
                         %% Then, before last author ...
                         %% \and
                         %% \author{<author name>
                         %% \affil{<number>}{}}

                         %% For example, assuming Garcia and Sonnery are both affiliated with
                         %% Universidad de Murcia:
                         %% \author{Roberta Graff\affil{1}{University of Cambridge, Cambridge,
                         %% United Kingdom},
                         %% Javier de Ruiz Garcia\affil{2}{Universidad de Murcia, Bioquimica y Biologia
                         %% Molecular, Murcia, Spain}, \and Franklin Sonnery\affil{2}{}}

                         \author{Carl Boettiger\affil{1}{Center for Population Biology, University of California, Davis}, 
                         Alan Hastings\affil{2}{Department of Environmental Science and Policy, University of California, Davis} }

                         \contributor{Submitted to Proceedings of the National Academy of Sciences
                         of the United States of America}

                         %% The \maketitle command is necessary to build the title page.
                         \maketitle

                         %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
                         \begin{article}

@@

                             %% When adding keywords, separate each term with a straight line: |
                             \keywords{ early warning signals | tipping point | alternative stable states | likelihood methods }

                             %% Optional for entering abbreviations, separate the abbreviation from
                             %% its definition with a comma, separate each pair with a semicolon:
                             %% for example:
                             %% \abbreviations{SAM, self-assembled monolayer; OTS,
                             %% octadecyltrichlorosilane}

                             % \abbreviations{}

                             %% The first letter of the article should be drop cap: \dropcap{}
                             %\dropcap{I}n this article we study the evolution of ''almost-sharp'' fronts

                             %% Enter the text of your article beginning here and ending before
                             %% \begin{acknowledgements}
                             %% Section head commands for your reference:
                             %% \section{}
                             %% \subsection{}
                             %% \subsubsection{}

\dropcap{T}here is an increasing recognition of the importance of regime shifts or critical transitions at a variety of scales in ecological systems~\cite{Holling1973, Wissel1984, Scheffer2001, Scheffer2009, Drake2010, Carpenter2011}⁠. 
Many important ecosystems may currently be threatened with collapse, including corals~\cite{Bellwood2004}, fisheries~\cite{Berkes2006}⁠, lakes~\cite{Carpenter2011}, and semi-arid ecosystems~\cite{Kefi2007}⁠. 
Given the potential impact of  these shifts on the sustainable delivery of ecosystem services
and the idea that management responses would be needed either to avoid an undesirable shift or else to adapt to novel conditions,
it is important to develop the ability to predict impending regime shifts based on early warning signs. 
With a good model of the system, detail-oriented approaches could be useful~\cite{Lenton2009},
but in cases where specific models are not available general approaches are needed~\cite{Scheffer2009}⁠.
We focus on the common scenario of a gradual environmental change yielding a sudden shift or bifurcation in a system.\footnote{A system involving a local bifurcation where the eigenvalue approaches zero \cite{Scheffer2001, Scheffer2009}.}  
Making this assumption provides a ``best-case'' test of detection approaches;
we will demonstrate the resulting uncertainty can be very high.
We describe a novel statistical approach for handling uncertainty that is appropriate for the goal of informing management.

Though previous research has identified patterns such as increasing variance or autocorrelation
in time-series approaching a critical transition,
there has been no method for forecasting the \emph{probability} of a collapse from a natural system.
Real-world applications of early warning signals will not be able to point to such differences between controlled and
manipulated systems that have been utilized in experimental studies of early warning~\cite{Drake2010, Carpenter2011}⁠.
While the number of potential patterns grows steadily~\cite{Carpenter2006, Dakos2008, Guttal2008, Guttal2008a, Dakos2011}, % add spectral methods
there are few examples that statistically quantify the patterns~\cite{Dakos2008, Dakos2011},⁠
and fewer still that statistically estimate the false alarm rate~\cite{Dakos2008} or 
the probability of missed detection⁠. 
We need to quantify such uncertainty and error rates to compare between possible methods~\cite{Contamin2009}
and address the continual concern about how much data is necessary⁠ for robust detection~\cite{Scheffer2001, Dakos2008, Carpenter2011, Scheffer2010, Inman2011}.

Detection of impending transitions is not just a scientific question 
and the goal should not solely be to detect with a 5\% chance of a false positive:
a higher false alarm rate may well be  worth a smaller chance of  missing a catastrophic shift. 
There is an obvious trade-off between the frequency with which a given approach will experience a false alarm and
with which it will fail to provide any advanced warning. 
Choosing the appropriate trade off requires knowledge of the likelihood of both kinds of error.
In this paper, we
(a) present the first measurements of reliability and sensitivity for early-warning indicators, 
(b) we use this method to estimate the amount of data required to achieve a given level of confidence,  
(c) we show where existing methods have performed little better than chance due to a lack of sufficient data or the use of insensitive indicators 
(d) we introduce a new indicator based on likelihood which substantially out-performs the existing methods.

We illustrate the trade-off between false alarms and failed detection using 
receiver-operating characteristic (ROC) curves first developed in signal-processing literature~\cite{Green1989, Keller2009}⁠. 
The curves represent the corresponding false alarm rate at any detection sensitivity (true positive rate), Fig 1.
The closer these distributions are to one-another, the more severe the trade-off.  
If the distributions overlap exactly, the ROC curve has a constant slope of unity.  
The ROC curve demonstrates this trade-off between accuracy and sensitivity.  
Different early-warning indicators will vary in their sensitivity to detect differences between stable systems and those approaching a critical transition, making the ROC curves a natural way to compare their performance.  
Since the shape of the curve will also depend on the duration and frequency of the time-series observations,
we can use these curves to illustrate by how much a given increase in sampling effort can decrease the rate of false alarms or failed detections.  
The area under the ROC curve provides a summary of how well the method compares to a random guess --
an area of 0.5 corresponding to perfectly overlapping distributions is performing no better than chance, while a statistic with area of unity always distinguishes correctly between stable systems and those approaching a transition.

To generate ROC curves for common early warning indicators such as 
increases in autocorrelation, variance, and skew, 
we must quantify the increase and generate the expected distributions of the focal statistic.  
While the existing literature frequently presents only a 
% e.g. Drake \& Griffen use any two-sigma deviation as indicative, yet any sufficiently sampled null process will produce a positive warning,  
% so this is not a reasonable statistical characterization
visual or heuristic measure of increase\footnote{That is, do not provide $p$-values or a statistic for which they could be generated.} in early-warning indicators~\cite{Scheffer2009, Drake2010, Carpenter2011, Carpenter2006}, 
we rely on the few examples that quantify an increasing trend using Kendall's $\tau$ correlation statistic~\cite{Dakos2008, Dakos2011, Dakos2009}.
Values of $\tau$ near unity indicate a strongly increasing trend in the warning indicator --
suggestive of an approaching transition, while values near zero suggest a lack of a trend -- characteristic of stable systems.
In principle, other quantifications of increasing variance, autocorrelation, etc.,
may perform better than $\tau$ and could be analyzed just as well in this framework. 
While we use this established metric of increase, we also introduce a novel warning indicator
that is not based on the gradual increase of a summary statistic but on likelihood comparisons (see Methods).

Having identified an indicator statistic,
we must be able to generate its distribution under both the scenario of a stable system and 
one approaching a transition before we can compute the ROC curve.
We fit general stochastic models representing common bifurcations
(e.g. saddle-node~\cite{Scheffer2009, Guttal2008a, VanNes2007, Biggs2009} and transcritical bifurcations~\cite{Drake2010})
for both scenarios to the data by maximum likelihood, and then simulate 500 replicate data sets under each⁠. 
By estimating the statistic (e.g. $\tau$) on each replicate we generate distributions such as depicted in Fig. 1 and compute ROC curves (see Methods).
Our approach provides a novel method for comparing alternatives by their likelihood ratio statistic~\cite{Cox1961}⁠ instead of $\tau$,
which provides a more direct and more powerful test than detecting an increase in summary statistics.
As these methods are computationally demanding, we provide a multi-processor enabled R package to assist with the analysis.

\section{Examples}

We illustrate this analysis of uncertainty in early warning indicators in simulated and empirical data sets (Fig. 2). 
We explore two empirical examples:
deuterium concentrations (Fig. 2c) provide an observational dataset previously cited as an early warning signal of a critical transition in glaciation~\cite{Dakos2008},
and \emph{Daphnia magna} concentrations (Fig. 2d) manipulated in a chemostat towards a critical transition~\cite{Drake2010}.  
For comparison, we include an individual-based simulation of a system approaching a saddle-node bifurcation due to environmental deterioration,
and the same system in a constant environment. 
The challenge of accurate detection is clearly illustrated by comparing the patterns for a system approaching transition (b-d) 
against the simulation of a stable system, in which chance fluctuations give the appearance of steadily rising autocorrelation. 
While such a pattern will fade in a longer or more frequently sampled time-series of the same system, 
we need a way to identify such cases where we lack the data to trust the pattern.

The ROC curves for these data (Fig. 3) show that the summary-statistic based indicators 
frequently lack the sensitivity to distinguish reliably between observed patterns from a stable or unstable system. 
The large correlations observed in the empirical examples (Fig. 2) are not uncommon in stable systems. 
For instance, we can read from the curve what fraction of true positives would be detected by a variance-based indicator
set to a 5\% false positive rate:
25\% in the simulated set,
5.0\% in the \emph{Daphnia} set,
and 5.4\% in the Glaciation data set.
Compare this to likelihood: 61\%, 34\%, and 100\% of the replicate true positives
would be correctly identified from the respective systems.
The value of the ROC curve is that we need not restrict the analysis to a pre-specified error such as 5\%,
as we would in classical hypothesis testing.  
We could instead decide that we wanted no to catch no less than 90\% of the true-positives,
and read off the corresponding false positives.  
For variance this corresponds respectively to a 49\%, 81\%, and 93\% false positive rate,
while for likelihood a 55\%, 35\%, 0\% false positive.\footnote{
Each false-positive, true-positive coordinate on the ROC curve corresponds to a threshold of the indicator statistic,
as illustrated in Fig. 1.  
Comparing this threshold with the observed value classifies the signal as detected or undetected.  See SI text.}
This flexibility is particularly valuable in the context of critical transitions, 
where data are often limited and the consequences of missed detection can be far more severe than false alarms.

\section{Discussion}
A pressing question in the application of early warning signals is quantifying
when the data are sufficient for detection to be reliable~\cite{Scheffer2009, Scheffer2010, Inman2011}. 
Our approach provides a natural way to address this challenge by estimating the ROC curve under differing data sampling efforts
for each of the indicators (Fig 4). 
Using our generic models estimated from the original data as in Fig. 3, 
we generate the corresponding distributions by simulating under the models as before, 
but with more or less frequent sampling to illustrate how our ability to detect an approaching transition 
would be hampered or improved with more or less data.  
While all methods improve as the number of data points increase, 
the coefficient of variation and the variance show more substantial improvement than autocorrelation, 
as expected~\cite{Carpenter2011};⁠
skew performs little better than chance, and likelihood is consistently the most reliable and sensitive (Fig. 4).

%% Threshold
%% Imagine a missed detection of no more than 10% is acceptable.

By assuming the underlying model corresponds to a saddle-node or transcritical bifurcation,
our analysis presents a ``best-case scenario'' for both the summary statistics and the likelihood-based approach. 
Other literature has already begun to address the additional challenges posed when the underlying 
dynamics do not correspond to these models~\cite{Hastings2010}.
Our results illustrate that even in this best-case scenario, 
reliable identification of warning signals from summary statistics can be difficult.  
The extent of this difficulty is not fully apparent from the warning statistic itself, Fig 2,
or even the distribution of that statistic under the null model of a stable system, 
but becomes clear from the ROC curve, where a 5\% false positive rate often corresponds to only a 5\% true positive rate as well.  
By estimating the ROC curve for a given set of data, 
we can better avoid applying warning signals in cases of inadequate power.
By taking advantage of the assumptions being made to write down a specific likelihood function,
we can develop indicators that get the most from this data.

%%% Model adequacy
In any application of early warning signals, it is essential to address the question of model adequacy.  
This is every bit as important for the summary-statistic based indicators such as variance and autocorrelation
as it is for the approach taken here; these statistics correspond to particular assumptions about the underlying process,
and systems that violate these assumptions can exhibit sudden transitions without any warning,
such as chaotic systems or those with non-smooth potential functions~\cite{Hastings2010}.
Sudden transitions have also been observed in ecological models in which 
the variance deceases rather then increases as the system nears a bifurcation point~\cite{Schreiber2003, Schreiber2008}.  
Our approach formalizes the assumptions about the underlying process to match the assumptions of the other warning signals.  
As the bifurcation results from the principle eigenvalue passing through zero, 
the warning signal is expected in linear-order dynamics;
estimation of the nonlinear model is less powerful and less accurate.  
The performance of this approach in the simulated data, which is nonlinear with non-Gaussian noise 
(introduced by the Poisson process demographic events), 
demonstrates the accuracy under violation of these assumptions.

                             %% == end of paper:

                             %% Optional Materials and Methods Section
                             %% The Materials and Methods section header will be added automatically.

                             %% Enter any subheads and the Materials and Methods text below.
                             \begin{materials}
                               \subsection{Methods}\label{methods}
For each dataset, parameters are estimated by maximum likelihood for a stable model
and a model of a system approaching a critical transition (Appendix).
The models are stochastic differential equations (SDEs)
that can be used to simulate data under the same sampling regime as the original data, 
or at any specified set of time points.
Distributions are determined by a parametric bootstrap:
500 replicate datasets are simulated under both stable and deteriorating stability model (for each dataset).
For each replicate, the summary statistics (variance, autocorrelation, skew, and coefficient of variation)
were computed over a sliding window equal to half the observed time (other window sizes had worse performance, see R package), 
and a Kendall's correlation test applied to estimate correlation coefficient $\tau$.
The estimates of $\tau$ under the replicate simulations from the stable system
and the estimates under the deteriorating stability system
make up the two distributions which are integrated at a sliding threshold to create the ROC curves, as illustrated in Fig 1.

The likelihood test is performed as in the parametric bootstrap; 
but each simulated replicate is refit to both stable and deteriorating models to generate a difference in likelihood scores~\cite{Cox1961}.
Refitting the models on each replicate, 
rather merely computing the likelihood under the original estimates,
makes this much more computationally intensive than the summary statistic calculation but is essential to the power of this approach~\cite{Huelsenbeck1996}.

When replicates are created by simulation under the estimated models,
we record data at the same time intervals observed in the original data (as in Fig. 3, main text). 
It is straight forward to allow this interval to be different,
representing the effect of increased or decreased sampling effort on the reliability-sensitivity trade-off in the ROC curve, Fig 4.

The source code for implementing each of the analyses and generating the plots presented is provided as an R package, \verb|warningsignals|.

                               \subsection{Data sets}
The simulated data sets are produced with 40 data points
sampled evenly sampled over a time interval (0,100) using an individual-based model of a system containing a saddle-node bifurcation, Eq~\eqref{master}.
The data was then fit to the LSN model, Eq~\eqref{LSN}, to compute the ROC curves.

The Glaciation data is accessible from NOAA~\cite{Petit1999}.
The data is preprocessed by linear interpolation and de-trending by Gaussian kernel smoothing 
to be as consistent as possible with the original analysis from Dakos \emph{et al.}~\cite{Dakos2008},
analyzing the third glaciation event, consisting of 121 sample points. 
The match is not exact since estimates the de-trending window size manually,
but the estimated correlations in the first-order auto-regression coefficients are in close agreement with that analysis. 
De-trending is intended to make the data consistent with the assumptions of the warning signal detection~\cite{Dakos2008}, 
which did not apply to the other data sets~\cite{Drake2010}.

The \emph{Daphnia} example comes from the chemostat ``H6'' in the experiments of Drake \& Griffen~\cite{Drake2010}. 
This individual replicates was chosen as an example of a single replicate 
that showed a statistically significant correlation in variance over window of time where critical slowing down was expected. 
Further empirical examples that have been previously investigated for early warning signs are also analyzed under our approach in SI Text.

All data and code for simulations and data processing are found in the R package.

                             \end{materials}

                             %% Optional Appendix or Appendices
                             \appendix 
                             \subsection{Models}
The detection of early warning is based on a \emph{local} theory,
predicting changes in the dynamics near the current attractor.
Consequently, it is as unnecessary as it is impossible to estimate a fully nonlinear model,
but only the linearized model near the bifurcation.

>>>>

%        File: ecologyletters.tex
%     Created: Sat Oct 29 02:00 PM 2011 P
% Last Change: Sat Oct 29 02:00 PM 2011 P
%

% Title page
% - author contributions
%- the article title
%- the full name(s) of all author(s), affiliation(s), e-mail address(es) of all author(s)
%- a short running title (abbreviated form of title) of less than 45 characters including spaces
%- up to 10 keywords for indexing purposes. It is very important that the keywords be chosen carefully.
%- the type of article (Ideas and Perspectives, Letter, or Reviews and Syntheses)
%- the number of words in the abstract, in the manuscript as a whole, and in the main text (excluding abstract, acknowledgements, references, table and figure legends)
%- the number of references (or for papers with extensive meta-analyses, the number of data-source references AND the number of general references)
%- the number of figures and tables
%- the name and complete mailing address (including telephone and fax numbers and e-mail address) of the person to whom correspondence should be sent.
%

%\documentclass[authoryear,5p]{elsarticle}
\documentclass[authoryear,preprint,11pt]{elsarticle}
\bibliographystyle{elsarticle-harv}
\usepackage{graphicx}
\usepackage{amsmath,amsfonts}
\usepackage{lineno}
\linenumbers
\usepackage{subfigure}
\usepackage[pdftex]{color}
\definecolor{darkblue}{rgb}{0,0,0.5}
\definecolor{darkgreen}{rgb}{0,0.5,0}
%\usepackage[pdftex, colorlinks, citecolor=darkblue,linkcolor=darkgreen]{hyperref}
\usepackage[pdftex, colorlinks]{hyperref}
\textwidth 6.75in
\oddsidemargin -0.15in
\evensidemargin -0.15in
\textheight 9in
\topmargin -0.5in
\newcommand{\ud}{\mathrm{d}}
\newcommand{\E}{\mathrm{E}}
\newcommand{\C}{\mathrm{Cov}}
\newcommand{\V}{\mathrm{Var}}

\journal{\tiny Ecology Letters} 
\begin{document}
\begin{frontmatter}
  \title{Quantifying Limits to Detection of Early Warning for Critical Transitions}
  \author[cpb]{Carl Boettiger\corref{cor1}}
  \ead{cboettig@ucdavis.edu, 610-389-6087}
  \author[esp]{Alan Hastings}
  %\author[info]{}
  %\author[davis]{}
  \cortext[cor1]{Corresponding author.}
  \address[cpb]{Center for Population Biology, 1 Shields Avenue, University of California, Davis, CA, 95616 United States.}
  \address[esp]{Department of Environmental Science and Policy, University of California, Davis}

 % \address[info]{ \\ 
  %              }

@@

  \begin{keyword}
early warning signals \sep tipping point \sep alternative stable states \sep likelihood methods 
   \end{keyword}
 \end{frontmatter}
\emph{\noindent \textbf{Running title}: Limits to Detection of Early Warning \\ 
  \textbf{Article Type}: Letter.  Abstract: 132 words, Text: 3,472 words, 34 references, 4 Figures. \\
  \textbf{Author Contributions}: C.B. wrote the software and performed the analysis with advice from A.H.  C.B. and A.H. designed the study and wrote the paper. 
}

\section{Introduction}
There is an increasing recognition of the importance of regime shifts or critical transitions at a variety of scales in ecological systems~\citep{Holling1973, Wissel1984, Scheffer2001, Scheffer2009, Drake2010, Carpenter2011}⁠. 
Many important ecosystems may currently be threatened with collapse, including corals~\citep{Bellwood2004}, fisheries~\citep{Berkes2006}⁠, lakes~\citep{Carpenter2011}, and semi-arid ecosystems~\citep{Kefi2007}⁠. 
Given the potential impact of  these shifts on the sustainable delivery of ecosystem services
and the idea that management responses would be needed either to avoid an undesirable shift or else to adapt to novel conditions,
it is important to develop the ability to predict impending regime shifts based on early warning signs. 
With a good model of the system, detail-oriented approaches could be useful~\citep{Lenton2009},
but in cases where specific models are not available general approaches are needed~\citep{Scheffer2009}⁠.

We focus on the common scenario of a gradual environmental change yielding a sudden shift or bifurcation in a system 
(\emph{i.e.} where the dominant eigenvalue passes through zero,
as discussed in~\citet{Scheffer2001, Scheffer2009}). 
Making this assumption provides a ``best-case'' test of detection approaches;
we will demonstrate the resulting uncertainty can be very high.
We describe a novel statistical approach for handling uncertainty that is appropriate for the goal of informing management.

\subsection{The implicit model}
Foundational work on early warning signals has operated under the often-implicit assumption that the system dynamics contain a saddle-node bifurcation and looks for patterns that would be associated with approaching this bifurcation point.  Such methods

\subsection{Explicit use of the model}
Detection of warning signals as model choice (``non-parametric'' Carpenter \& Brock, unpublished Dakos, Kuehn arxiv, Dakos 2011?, Gross 2011)

\subsection{Why information criteria will not serve.}

\subsection{Beyond hypothesis testing}
Hypothesis testing is built upon the objective of

\citet{Seekell2011}

\subsection{Visualizing the informativeness through ROC curves}

%% EDIT -- 
Though previous research has identified patterns such as increasing variance or autocorrelation
in time-series approaching a critical transition,
there has been no method for forecasting the \emph{probability} of a collapse from a natural system.
Real-world applications of early warning signals will not be able to point to such differences between controlled and
manipulated systems that have been utilized in experimental studies of early warning~\citep{Drake2010, Carpenter2011}⁠.
While the number of potential patterns grows steadily~\citep{Carpenter2006, Dakos2008, Guttal2008, Guttal2008a, Dakos2011}, % add spectral methods
there are few examples that statistically quantify the patterns~\citep{Dakos2008, Dakos2011},⁠
and fewer still that statistically estimate the false alarm rate~\citep{Dakos2008} or 
the probability of missed detection⁠. 
We need to quantify such uncertainty and error rates to compare between possible methods~\citep{Contamin2009}
and address the continual concern about how much data is necessary⁠ for robust detection~\citep{Scheffer2001, Dakos2008, Carpenter2011, Scheffer2010, Inman2011}.

%% EDIT
Detection of impending transitions is not just a scientific question 
and the goal should not solely be to detect with a 5\% chance of a false positive:
a higher false alarm rate may well be  worth a smaller chance of  missing a catastrophic shift. 
There is an obvious trade-off between the frequency with which a given approach will experience a false alarm and
with which it will fail to provide any advanced warning. 
Choosing the appropriate trade off requires knowledge of the likelihood of both kinds of error.
In this paper, we
(a) present the first measurements of reliability and sensitivity for early-warning indicators, 
(b) we use this method to estimate the amount of data required to achieve a given level of confidence,  
(c) we show where existing methods have performed little better than chance due to a lack of sufficient data or the use of insensitive indicators 
(d) we introduce a new indicator based on likelihood which substantially out-performs the existing methods.

We illustrate the trade-off between false alarms and failed detection using 
receiver-operating characteristic (ROC) curves first developed in signal-processing literature~\citep{Green1989, Keller2009}⁠. 
The curves represent the corresponding false alarm rate at any detection sensitivity (true positive rate), Fig 1.
The closer these distributions are to one-another, the more severe the trade-off.  
If the distributions overlap exactly, the ROC curve has a constant slope of unity.  
The ROC curve demonstrates this trade-off between accuracy and sensitivity.  
Different early-warning indicators will vary in their sensitivity to detect differences between stable systems and those approaching a critical transition, making the ROC curves a natural way to compare their performance.  
Since the shape of the curve will also depend on the duration and frequency of the time-series observations,
we can use these curves to illustrate by how much a given increase in sampling effort can decrease the rate of false alarms or failed detections.  
The area under the ROC curve provides a summary of how well the method compares to a random guess --
an area of 0.5 corresponding to perfectly overlapping distributions is performing no better than chance, while a statistic with area of unity always distinguishes correctly between stable systems and those approaching a transition.

To generate ROC curves for common early warning indicators such as 
increases in autocorrelation, variance, and skew, 
we must quantify the increase and generate the expected distributions of the focal statistic.  
While the existing literature frequently presents only a 
% e.g. Drake \& Griffen use any two-sigma deviation as indicative, yet any sufficiently sampled null process will produce a positive warning,  
% so this is not a reasonable statistical characterization
visual or heuristic measure of increase (\emph{i.e.}, do not provide $p$-values or a statistic for which they could be generated) in early-warning indicators~\citep{Scheffer2009, Drake2010, Carpenter2011, Carpenter2006}, 
we rely on the few examples that quantify an increasing trend using Kendall's $\tau$ correlation statistic~\citep{Dakos2008, Dakos2011, Dakos2009}.
Values of $\tau$ near unity indicate a strongly increasing trend in the warning indicator --
suggestive of an approaching transition, while values near zero suggest a lack of a trend -- characteristic of stable systems.
In principle, other quantifications of increasing variance, autocorrelation, etc.,
may perform better than $\tau$ and could be analyzed just as well in this framework. 
While we use this established metric of increase, we also introduce a novel warning indicator
that is not based on the gradual increase of a summary statistic but on likelihood comparisons (see Methods).

Having identified an indicator statistic,
we must be able to generate its distribution under both the scenario of a stable system and 
one approaching a transition before we can compute the ROC curve.
We fit general stochastic models representing common bifurcations
(\emph{e.g.} saddle-node~\citep{Scheffer2009, Guttal2008a, VanNes2007, Biggs2009} and transcritical bifurcations~\citep{Drake2010})
for both scenarios to the data by maximum likelihood, and then simulate 500 replicate data sets under each⁠. 
By estimating the statistic (\emph{e.g.} $\tau$) on each replicate we generate distributions such as depicted in Fig. 1 and compute ROC curves (see Methods).
Our approach provides a novel method for comparing alternatives by their likelihood ratio statistic~\citep{Cox1961}⁠ instead of $\tau$,
which provides a more direct and more powerful test than detecting an increase in summary statistics.
As these methods are computationally demanding, we provide a multi-processor enabled R package to assist with the analysis.

\section{Materials \& Methods}\label{methods}

\subsection{Statistical methodology}
For each given time-series, parameters are estimated by maximum likelihood 
for a stable model and a model of a system approaching a critical transition (see Models).
The models are stochastic differential equations (SDEs)
that can be used to simulate data under the same sampling regime as the original data, 
or at any specified set of time points.
Distributions are determined by a parametric bootstrap:
500 replicate datasets are simulated under both stable and deteriorating stability model (for each dataset).
For each replicate, the summary statistics (variance, autocorrelation, skew, and coefficient of variation)
were computed over a sliding window equal to half the observed time (other window sizes had worse performance, see R package), 
and a Kendall's correlation test applied to estimate correlation coefficient $\tau$.
The estimates of $\tau$ under the replicate simulations from the stable system
and the estimates under the deteriorating stability system
make up the two distributions which are integrated at a sliding threshold to create the ROC curves, as illustrated in Fig 1.

The likelihood test is performed as in the parametric bootstrap; 
but each simulated replicate is refit to both stable and deteriorating models to generate a difference in likelihood scores~\citep{Cox1961}.
Refitting the models on each replicate, 
rather merely computing the likelihood under the original estimates,
makes this much more computationally intensive than the summary statistic calculation but is essential to the power of this approach~\citep{Huelsenbeck1996}.
When replicates are created by simulation under the estimated models,
we record data at the same time intervals observed in the original data (as in Fig. 3, main text). 
It is straight forward to allow this interval to be different,
representing the effect of increased or decreased sampling effort on the reliability-sensitivity trade-off in the ROC curve, Fig 4.
The source code for implementing each of the analyses and generating the plots presented is provided as an R package, \verb|warningsignals|.

\subsection{Example Data}
The simulated data sets are produced with 40 data points
sampled evenly sampled over a time interval (0,100) using an individual-based model of a system containing a saddle-node bifurcation, Eq~\eqref{master}.
The data was then fit to the LSN model, Eq~\eqref{LSN}, to compute the ROC curves.

The Glaciation data is accessible from NOAA~\citep{Petit1999}.
The data is preprocessed by linear interpolation and de-trending by Gaussian kernel smoothing 
to be as consistent as possible with the original analysis from Dakos \emph{et al.}~\citep{Dakos2008},
analyzing the third glaciation event, consisting of 121 sample points. 
The match is not exact since estimates the de-trending window size manually,
but the estimated correlations in the first-order auto-regression coefficients are in close agreement with that analysis. 
De-trending is intended to make the data consistent with the assumptions of the warning signal detection~\citep{Dakos2008}, 
which did not apply to the other data sets~\citep{Drake2010}.

The \emph{Daphnia} example comes from the chemostat ``H6'' in the experiments of Drake \& Griffen~\citep{Drake2010}. 
This individual replicates was chosen as an example of a single replicate 
that showed a statistically significant correlation in variance over window of time where critical slowing down was expected.  All data and code for simulations and data processing are found in the R package.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Supp. materials or main text?  %
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Models}
The detection of early warning is based on a \emph{local} theory,
predicting changes in the dynamics near the current attractor.
Consequently, it is as unnecessary as it is impossible to estimate a fully nonlinear model,
but we can approximate the model near the bifurcation.  
Many possible complex models will correspond to this approximation.

%\textbf{Saddle-node bifurcation.}
One of the most common critical transitions modeled in ecological systems with alternative stable states is a saddle-node bifurcation~\citep[\{emph{e.g.}[]{Scheffer2001, Scheffer2009, Guttal2008a, VanNes2007, Biggs2009}.
The individual-based simulation follows the master equation, 
\begin{align}
  P(n,t) &= b_{n-1} p_{n-1} + d_{n+1}p_{n+1} - (b_n+d_n)p_n  \label{master} \\ 
  b_n &= \frac{e K x^2}{X^2 + h_t^2} \\ 
  d_n &= e X_t - a_t   
\end{align}
This model can experience a saddle-node bifurcation through increased mortality by increasing either $e$ or $a$. 
The SDE approximation holds for reasonable population sizes:
\begin{equation}
dX_t = \left( \frac{e K x^2}{X^2 + h_t^2} - e X_t - a_t\right) dt + \sigma \sqrt{ \frac{e K x^2}{X^2 + h_t^2} + e X_t + a_t} dB_t \label{ass}
\end{equation}
This parameterization is only an example, of course many others are possible~\citep{Scheffer2009, Scheffer2001, Strogatz2001a, Guckenheimer1983}. 
Under appropriate transformation, all can be written in the same normal form for he saddle-node bifurcation,
\begin{equation}
\frac{\ud x}{\ud t} = r_t- x^2.
\label{saddle-node}
\end{equation}
While in principle this model could be estimated directly from the data,
if all observations have been prior to the crash (\emph{i.e.} near the desirable equilibrium)
we have no data which is informative about the nonlinearity and such estimates will be unreliable.  
Fortunately the theory of early warning signals is a \emph{local} one, 
and we can linearize around the equilibrium to obtain a model we can estimate both more easily and more accurately.
